XSGRB - The X-shooter GRB afterglow sample

### 2017-01-03 document/XSGRB_sample.tex
NH distribution

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

@@

@@

@@

>>>>

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

@@

Because we only follow up targets with either a clear afterglow or counterpart, correctly associating a galaxy with a burst when there is no optical afterglow is more difficult.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_dist.pdf}
	\caption{Distributions of DLA hydrogen column densities for DLAs found in
	quasar absorption lines, from \citep{Noterdaeme2012b} in blue. Overplot in green is the kernel density estimate
	is the similar distribution, only for DLAs in GRB sightlines. Values are taken
	from the compilation in Tanvir2017 et al. (in prep)\citet{Tanvir2017}, along
	with the new values presented in this sample. The marginal distributions for
	the two distribution are also shown, where the different environments probed
	are clearly visible in the hydrogen column densities, as previously also noted
	in \citet{Fynbo2009}}
	\label{fig:NH_dist}
\end{figure}

@@

In a similar manner to how the X-ray derived column densities was used to assess
the sample completeness in Sect. \ref{completeness}, we addresses the same
question in terms of the optically derived hydrogen column densities. Using the
compilation by \citet{Tanvir2017} we have all 81 published HI values previously,
we can compare to the 39 new HI columns added by this sample. We note that this
sample increase the number of optically derived hydrogen column densities with
$\sim$ 50 percent. We compare the median, the 14th, and 86th percentiles of the
two distributions where the sample presented here has $21.8_{0.8}^{0.3}$ and the
rest of the litterature values has $21.5_{1.5}^{0.4}$. We see that the two
distribution has a large degree of overlap due to the large width of the
distributions, but find a slightly higher median value for the new sample
presented here. A 2-sided KS test gets us a p-value of p = 0.006, meaning
relatively large evidence against the null that the two samples are drawn from
the same underlying distribution. Because the bursts that have measurements of
the hydrogen column density are selected sole based on our ability to infer a
column, it is difficult to make any strong conclusions about the completeness in
terms of gas content. We show the comparison between the two samples in Fig.
\ref{fig:NH_dist}, where we additionally show the column-distribution for 12081
quasar DLAs published in \citet{Noterdaeme2012b}. The ability of the GRBs to
select the highest N$_H$ is immediately visible as is also noted previously
\citep{Prochaska2007, Fynbo2009}. The reason for this is the selection of sight
lines as random position of the QSO DLAs whereas the GRB forms in dense regions
in the host. The N$_H$ distribution has also be used the infer the position of
the gas in host \cite{Buchner2016} where a galaxy-wide gas reservoir is
preferred.
